---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# skrunchy

<!-- badges: start -->
<!-- badges: end -->

 **Sk**eena River **Run** Reconstruction for **Ch**inook Salmon


The goal of skrunchy is to recreate and update the run reconstruction for Skeena River (summer run timing) Chinook upstream of Tyee test fishery, as documented in [Winther et al. 2024](https://publications.gc.ca/site/eng/9.901355/publication.html "An assessment of Skeena River Chinook salmon using genetic stock identification 1984 to 2020")

Detailed methods with variables and equations matching (as much as possible) this package will be added in the file: `methods.PDF`. 

## General methods

Here, we take a known abundance of population $K$ and expand it to estimate an aggregate population size $X_{aggregate}$ using the proportion of fish from population $K$ in a mixed genetic sample, $P_K$: 

$$X_{aggregate} = \frac{K}{P_K}$$

Further, the abundance of other populations $X_i$ can be estimated using their genetic proportion $P_i$ and the aggregate abundance $X_{aggregate}$: 

$$X_i = X_{aggregate} \cdot P_i$$

After that, the run is "reconstructed", i.e., from the time the fish are approximately age 3 ocean fish, all mortalities (fishery, incidental mortality, brood stock) are added to the estimate of spawners to estimate total recruits produced by each cohort. This produces estimates of spawners and recruits by brood year, which can then be use to model productivity. 


## Installation

You can install the development version of skrunchy from [GitHub](https://github.com/) with:

```
install.packages("pak")
pak::pak("Pacific-salmon-assess/skrunchy")
```

```{r load-libraries, include = FALSE, echo = FALSE }
library(skrunchy)
library(ggplot2)
library(latex2exp)
```

## Examples (currently use fake data for several data sources - not accurate)

Make some **fake** genetic proportion and catch data (by week), expand to estimates of annual proportions, and plot:

```{r example_P_tilde, dpi = 300}

d <- make_P_G()
res <- get_P_tilde(P = d$P, sigma_P = d$sigma_P, G = d$G)
P_tilde <- res$P_tilde
dp <- res$df_merged

ggplot(dp, aes(y = P_tilde, x = y, group = i)) +
  geom_errorbar( aes( ymin = P_tilde - sigma_P_tilde, ymax = P_tilde + sigma_P_tilde ), colour="dodgerblue") +
  geom_point() +
  geom_line() +
  facet_wrap(~i, ncol=2) +
  ylab(TeX("$\\tilde{P}$")) +
  geom_hline(aes(yintercept=0)) +
  theme_classic()
  
```

Now do expansions to get returns to Terrace for each population plus the aggregate

```{r example_X, dpi = 300 }
k <- read.csv(here("data/kitsumkalum-escapement.csv"))
X <- get_X(P_tilde = res$P_tilde, sigma_P_tilde = res$sigma_P_tilde, K= k$kitsumkalum_escapement, 
           sigma_K = k$sd,
           y = k$year)
dt <- X$df_merged

ggplot(dt, aes(y = X, x = y, group = i)) +
  geom_errorbar( aes( ymin = X - sigma_X, ymax = X + sigma_X)) +
  geom_point() +
  geom_line() +
  facet_wrap(~i, ncol=2, scales = "free_y") +
  geom_hline(aes(yintercept=0)) +
  theme_classic()
  
```
Get escapement for each population, plot with returns to Terrace (note, will only be different for Skeena aggregate and the three upper populations)

```{r example_E, dpi = 300 }
Tau_U <- sample(5000:10000, size=length(k$y), replace=TRUE)
E <- get_E(K = k$kitsumkalum_escapement, X = X$X, Tau_U = Tau_U,
     known_population = "Kitsumkalum",
     aggregate_population = "Skeena",
     lower_populations = c("Lower Skeena", "Zymoetz-Fiddler"),
     upper_populations = c("Upper Skeena", "Middle Skeena", "Large Lakes"))
de <- E$df


ggplot(dt, aes(y = X, x = y, group = i)) +
  geom_errorbar( aes( ymin = X - sigma_X, ymax = X + sigma_X)) +
  geom_point() +
  geom_line() +
  geom_line(data = de, aes(y = E, x = y, group=i), colour="gray") +
  facet_wrap(~i, ncol=2, scales = "free_y") +
  ylab("Return to Terrace (X) in black, escapement (E) in gray") +
  geom_hline(aes(yintercept=0)) +
  theme_classic()

```

Get age proportions by age, population and year (fake data).


```{r example_omega, dpi = 300 }
 populations <- c("Kitsumkalum", "Lower Skeena", "Zymoetz-Fiddler", "Upper Skeena", "Middle Skeena", "Large Lakes", "Skeena")
  n_populations <- length(populations)
  years <- k$year
  n_years <- length(years)
  ages <- c(4,5,6,7)
  p_ages <- c(30,40,40,1) # make up relative frequency of different ages
  n_ages <- length(ages)
  # Make up some age data - number of aged fish of each age
  d <- sapply(p_ages, FUN = function(x){ rpois( n = n_populations*n_years, lambda= x) })
  n <- array( d,  dim = c(n_populations, n_years, n_ages), dimnames = list(i = populations, y = years, a = ages))
  omega <- get_omega(n)
do <- omega$df

ggplot( do, aes(y = omega, x = y, group = i)) +
  geom_point() + 
  geom_line() + 
  geom_hline(aes(yintercept=0)) + 
  ylab(TeX("$\\Omega$") )+
  facet_grid( i ~ a ) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))


```

Get age-specific escapement by using age proportions. 

```{r example_E_star, dpi=300}
E_star <- get_E_star(E = E$E, omega = omega$omega)

ggplot( E_star$df, aes(y = E_star, x = y, group = i)) +
  geom_point( colour="gray") + 
  geom_line( colour="gray") + 
  geom_hline(aes(yintercept=0)) + 
  facet_grid( i ~ a , scales = "free_y") + 
  ylab("Escapement (E*)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Get spawners for each population (accounts for brood stock removals), plot with returns to Terrace and escapement. Spawners should only be different from escapement for Skeena aggregate and Kitsumkalum, since brood removals are only for Kitsumkalum.

```{r example_S_star, dpi = 300 }
# make up brood removal data. Much larger than actual values, so that it is visible on plots. 
B_star <- E_star$E_star["Kitsumkalum",,] * runif(n = length(E_star$E_star["Kitsumkalum",,] ), 0.1, 0.2 )
B_star[,4] <- 0 # make age 7 brood = 0 so you don't get negative fish in S_star
S_star <- get_S_star(E_star = E_star$E_star, B_star = B_star)

ggplot( E_star$df, aes(y = E_star, x = y, group = i)) +
  geom_point(colour="gray") + 
  geom_line(colour="gray") + 
  geom_line(data = S_star$df, aes(y = S_star, x = y, group = i), colour="dodgerblue") +
  geom_hline(aes(yintercept=0)) + 
  facet_grid( i ~ a , scales = "free_y") + 
  ylab("Escapement (E*) in gray and spawners (S*) in blue") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))

```

Get wild spawners for each population (accounts for hatchery origin spawners), plot with returns to Terrace, escapement, and spawners. Wild spawners should only be different from spawners for Skeena aggregate and Kitsumkalum, since hatchery origin spawners only occur for Kitsumkalum.


```{r example_W_star, dpi = 300 }
# make up hatchery spawners data. Larger than actual values, so that it is visible on plots. 
H_star <- S_star$S_star["Kitsumkalum",,] * runif(n = length(S_star$S_star["Kitsumkalum",,] ), 0.1, 0.5 )
H_star[,4] <- 0 # make age 7 hatchery 0 to avoid negative wild spawner values.
## get Wild spawners
W_star <- get_W_star( S_star = S_star$S_star, H_star = H_star) 

ggplot( E_star$df, aes(y = E_star, x = y, group = i)) +
  geom_point(colour="gray") + 
  geom_line(colour="gray") + 
  geom_line(data = S_star$df, aes(y = S_star, x = y, group = i), colour="dodgerblue") +
  geom_line(data = W_star$df, aes(y = W_star, x = y, group = i), colour="firebrick") +
  geom_hline(aes(yintercept=0)) + 
  facet_grid( i ~ a , scales = "free_y") + 
  ylab("Escapement (E*) in gray, spawners (S*) in blue\n, and wild spawners (W*) in red.") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))

```

Get proportion wild spawners for each population and plot. Should only be <1 for Skeena aggregate and Kitsumkalum, since hatchery origin spawners only occur for Kitsumkalum. Note this is not real data, p is very high for Kitsumkalum across years. 


```{r example_p, dpi = 300 }
p <- get_p(W_star = W_star$W_star, E_star = E_star$E_star)

ggplot(p$df, aes(y = p, x = y, group = i)) +
  geom_point() +
  geom_line() +
  facet_grid(i~a) +
  ylab("Proportion wild spawners, p") +
  geom_hline(aes(yintercept=0)) +
  theme_classic()

```


Estimate terminal total mortalities in the lower Skeena by population, year, and age. 


```{r example_tau_L, dpi = 300 }
Tau_L <- sample(1000:2000, size = length(k$year))
tau_L <- get_tau_L(Tau_L = Tau_L, omega = omega$omega, P_tilde = P_tilde, aggregate_population = "Skeena")
dtauL <- tau_L$df 

ggplot( dtauL, aes(y =tau_L, x = y, group = i)) +
  geom_point() + 
  geom_line() + 
  geom_hline(aes(yintercept=0)) + 
  ylab(TeX("$\\tau_L$")) +
  facet_grid( i ~ a, scales="free_y") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Estimate terminal total mortalities in the upper Skeena by population, year, and age. Should be 0 for Kitsumkalum, Lower Skeena, and Zymoetz-Fiddler. 


```{r example_tau_U, dpi = 300 }
tau_U <- get_tau_U(Tau_U = Tau_U, omega = omega$omega, P_tilde = P_tilde, aggregate_population = "Skeena",
                   upper_populations = c("Middle Skeena", "Large Lakes", "Upper Skeena"),
    lower_populations = c("Lower Skeena", "Kitsumkalum", "Zymoetz-Fiddler") )

dtauU <- tau_U$df 

ggplot( dtauU, aes(y =tau_U, x = y, group = i)) +
  geom_point() + 
  geom_line() + 
  geom_hline(aes(yintercept=0)) + 
  ylab(TeX("$\\tau_U$")) +
  facet_grid( i ~ a) + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Estimate terminal total mortalities in the marine area by population, year, and age. 


```{r example_tau_M, dpi = 300 }
use_arr <- W_star$W_star["Kitsumkalum",,]
tau_dot_M <- array(runif(length(use_arr), 0.01, 0.2), dim = dim(use_arr), dimnames = dimnames(use_arr))
tau_M <- get_tau_M( W_star = W_star$W_star, tau_dot_M = tau_dot_M)

ggplot( tau_M$df, aes(y =tau_M, x = y, group = i)) +
  geom_point() + 
  geom_line() + 
  geom_hline(aes(yintercept=0)) + 
  ylab(TeX("$\\tau_M$")) +
  facet_grid( i ~ a, scales="free_y") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))
```

Get total terminal mortality 

```{r example_tau, dpi=300}
tau <- get_tau(tau_U = tau_U$tau_U, tau_L = tau_L$tau_L, tau_M= tau_M$tau_M)

ggplot( tau$df, aes(y =tau, x = y, group = i)) +
  geom_line() +
  geom_point() +
  geom_line( data = tau_L$df, aes(y =tau_L, x = y), colour="goldenrod") +
  geom_line( data = tau_U$df, aes(y =tau_U, x = y), colour="aquamarine") +
  geom_line( data = tau_M$df, aes(y =tau_M, x = y), colour="blue") +
  geom_hline(aes(yintercept=0)) + 
  ylab(TeX("$\\tau$")) +
  facet_grid( i ~ a, scales="free_y") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle=90, vjust=0.5))

```


Get wild total terminal mortality 

```{r example_tau_W, dpi=300}
```


Get wild terminal run 

```{r example_TermRun, dpi=300}
```

Get mature run 

```{r example_MatureRun, dpi=300}
```


